A pipeline to analyse time-course gene expression data

### 2020-07-15 scripts/manuscript.Rmd
start updating manuscript

**EAP**: Why is this list not the same as the (hidden via echo=FALSE) list at
the beginning of the document? I also needed to install `BiocWorkflowTools`,
`KEGGprofile` to run the .Rmd
**NV**: I think some of the libraries above are not needed anymore. Some are
also only used in hidden elements. I just checked that the whole pipeline
works, by executing all command lines the reader will see, and the libraries
are all loaded fine. I think BiocWorkflowTools is only need in the compilation
of the Rmd file.

@@

In both
situations, we first need to set up a object (a `moanin_model` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `moanin_model` object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the `moanin_model` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatement effect, and a
second one named `Timepoint` containing the timepoint information.

@@

We create a `moanin_model` for our data:

```{r}
moanin_model = create_moanin_model(meta, degrees_of_freedom=6)
```

@@

```{r basis_matrix}
print(moanin_model)
```

@@

```{r make_timepoints_contarts}
# Define contrasts	
contrasts = create_timepoints_contrasts("M", "VL", moanin_model)
```

@@

```{r weekly_differential_analysis}
weekly_de_analysis = DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)
```

@@

```{r echo=FALSE, results="hide"}
contrasts = create_timepoints_contrasts("M", "K", moanin_model)
K_weekly_de_analysis =  DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "C", moanin_model)
C_weekly_de_analysis =  DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts("M", "VH", moanin_model)
VH_weekly_de_analysis =  DE_timepoints(
     data, moanin_model, contrasts,
     use_voom_weights=FALSE)

@@

```{r timeSignifPerTime,fig.width=6, fig.height=6, echo=FALSE, fig.cap="Number of genes found DE in each timepoint when comparing the influenza strains to control."}
timely_barplot = function(de_analysis, 
    labels=NULL, qval_threshold=0.05,
    xlab="Timepoint", ylab="Number of DE genes", main="", ylim=c(0, 20000), col="black", ...){

    qval_colnames = colnames(de_analysis)[
	grepl("qval", colnames(de_analysis))]
    if(is.null(labels)){
	labels = sapply(
	    strsplit(gsub("-qval", "", qval_colnames), "\\."), .subset2, 3)
    }
    number_de_genes_per_time = colSums(de_analysis[, qval_colnames] < 0.05)

    barplot(number_de_genes_per_time, 
            names.arg=labels, xlab=xlab,ylab=ylab,
            col=col, main=main, ylim=ylim,...)  
}

par(mfrow=c(2, 2))
time = names(time_colors)
timely_barplot(K_weekly_de_analysis, labels=time, main="Kawasaki", las=3)
timely_barplot(C_weekly_de_analysis, labels=time, main="California", las=3)
timely_barplot(weekly_de_analysis, labels=time, main="Vietnam low-dose", las=3)
timely_barplot(VH_weekly_de_analysis, labels=time, main="Vietnam high-dose", las=3)
```

@@

```{r showMissingDEGenes, echo=FALSE, fig.cap="The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint."}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(data, moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group, smooth=TRUE)
```

@@

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
DE_results = DE_timecourse(
    data, moanin_model, timecourse_contrasts,
    use_voom_weights=FALSE)
```

@@

Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then under-expressed at the end of the
experiment. As a result, `moanin` provides a number of possible ways to
compute the log fold change across the whole time-course. This is done via the function `estimate_log_fold_change` which takes as
arguments the data, the `moanin_model` object, the contrasts to evaluate, and the method to use to estimate the log-fold change.

@@

```{r}
log_fold_change_timepoints = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts,  method="timely")
```

@@

```{r}
log_fold_change_timecourse = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts,  method="timecourse")

log_fold_change_sum = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts,  method="sum")

log_fold_change_max = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts, method="max")

log_fold_change_min = estimate_log_fold_change(
    data, moanin_model, timecourse_contrasts, method="min")
```

@@

```{r topPvalueGenes, fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(data[top_DE_genes_pval, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r topLogFold,fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(data[top_DE_genes_lfc, ], moanin_model,
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

- **Splines estimation**: for each gene, fit the splines function with the basis
  of your choice (as contained in the `moanin_model` object).
- **Rescaling splines**: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1 and thus comparable between genes.
- **K-means**: apply k-means on the rescaled fitted values of the splines to estimate the
  cluster centroids.

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(
    y, moanin_model, n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_splines_data(
    kmeans_clusters$centroids, moanin_model,
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

```{r genesVsCentroids, fig.height=6, fig.width=6}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(dataToPlot, moanin_model,
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r assigning_scores_and_labels}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    data, kmeans_clusters)
```

>>>>

@@

In both
situations, we first need to set up a object (a `Moanin` object) to
hold the meta data, as well as information for fitting the spline model
(formula, number of degrees of freedom of the splines, ...). The `Moanin` object
will contain information used
throughout this analysis, in particular the condition and timepoints of each sample and the
the basis matrix for fitting splines models.

We start by creating the `Moanin` object using the `create_moanin_model`
function. We need to provide two things to the function: a `data.frame` with
the metadata, and the number of degrees of freedom of the splines to be used in the
functional modeling. The metadata `data.frame` object should contain at least
two columns: one named `Group`, containing the  treatement effect, and a
second one named `Timepoint` containing the timepoint information.

@@

We create a `Moanin` object for our data:

```{r}
moanin_model = create_moanin_model(data=data, meta=meta,
    degrees_of_freedom=6)
```

@@

```{r basis_matrix}
show(moanin_model)
```

@@

```{r make_timepoints_contarts}
# Define contrasts	
contrasts = create_timepoints_contrasts(moanin_model,"M", "VL")
```

@@

```{r weekly_differential_analysis}
weekly_de_analysis = DE_timepoints( moanin_model, contrasts,
     use_voom_weights=FALSE)
```

@@

```{r echo=FALSE, results="hide"}
contrasts = create_timepoints_contrasts(moanin_model,"M", "K")
K_weekly_de_analysis =  DE_timepoints(
     moanin_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts(moanin_model,"M", "C")
C_weekly_de_analysis =  DE_timepoints(
     moanin_model, contrasts,
     use_voom_weights=FALSE)

contrasts = create_timepoints_contrasts( moanin_model,"M", "VH")
VH_weekly_de_analysis =  DE_timepoints(
     moanin_model, contrasts,
     use_voom_weights=FALSE)

@@

```{r timeSignifPerTime,fig.width=6, fig.height=6, echo=FALSE, fig.cap="Number of genes found DE in each timepoint when comparing the influenza strains to control."}
par(mfrow=c(2, 2))
time = names(time_colors)
perWeek_barplot(K_weekly_de_analysis, labels=time, main="Kawasaki", las=3)
perWeek_barplot(C_weekly_de_analysis, labels=time, main="California", las=3)
perWeek_barplot(weekly_de_analysis, labels=time, main="Vietnam low-dose", las=3)
perWeek_barplot(VH_weekly_de_analysis, labels=time, main="Vietnam high-dose", las=3)
```

@@

```{r showMissingDEGenes, echo=FALSE, fig.cap="The expression values over time for 10 genes found significant DE in timepoints 48,60,120,168H, but not in the 72H timepoint."}
exampleGenes<-names(signifCombos[signifCombos=="48,60,120,168"][1:10])
plot_splines_data(moanin_model, subset_data=exampleGenes,
		  colors=ann_colors$Group, smooth=TRUE)
```

@@

# The function takes the data (data.frame or named matrix), the meta data
# (data.frame containing a timepoint and group column, the first corresponding
# to the time-course information, the latter corresponding to the
# treatment).
DE_results = DE_timecourse( moanin_model, timecourse_contrasts,
    use_voom_weights=FALSE)
```

@@

Computing the log fold change on a time-course experiment is not trivial: one
can be interested in the average log-fold change across time, or the
cumulative log-fold change. Sometimes a gene can be over-expressed at the
beginning of the time-course data, and then under-expressed at the end of the
experiment. As a result, `moanin` provides a number of possible ways to
compute the log fold change across the whole time-course. This is done via the function `estimate_log_fold_change` which takes as
arguments the data, the `Moanin` object, the contrasts to evaluate, and the method to use to estimate the log-fold change.

@@

```{r}
log_fold_change_timepoints = estimate_log_fold_change(
    moanin_model, timecourse_contrasts,  method="timely")
```

@@

```{r}
log_fold_change_timecourse = estimate_log_fold_change(
    moanin_model, timecourse_contrasts,  method="timecourse")

log_fold_change_sum = estimate_log_fold_change(
    moanin_model, timecourse_contrasts,  method="sum")

log_fold_change_max = estimate_log_fold_change(
    moanin_model, timecourse_contrasts, method="max")

log_fold_change_min = estimate_log_fold_change(
    moanin_model, timecourse_contrasts, method="min")
```

@@

```{r topPvalueGenes, fig.height=6, fig.width=6}
top_DE_genes_pval = names(sort(pvalue)[1:10])
plot_splines_data(moanin_model,subset_data=top_DE_genes_pval, 
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r topLogFold,fig.height=6, fig.width=6}
top_DE_genes_lfc = names(
    sort(abs(lfc_timecourse),
	 decreasing=TRUE)[1:10])
plot_splines_data(moanin_model, subset_data=top_DE_genes_lfc, 
    colors=ann_colors$Group, smooth=TRUE,
    mar=c(1.5,2.5,2,0.1))

@@

- **Splines estimation**: for each gene, fit the splines function with the basis
  of your choice (as contained in the `Moanin` object).
- **Rescaling splines**: for each gene, rescale the estimated splines function
  such that the values are bounded between 0 and 1 and thus comparable between genes.
- **K-means**: apply k-means on the rescaled fitted values of the splines to estimate the
  cluster centroids.

@@

```{r clustering}
# First fit the kmeans clusters
kmeans_clusters = splines_kmeans(
    moanin_model[genes_to_keep, ], n_clusters=8,
    random_seed=42,
    n_init=20)
```

@@

```{r fig.height=6, fig.width=6}
plot_splines_data(moanin_model,
    data=kmeans_clusters$centroids, 
    colors=ann_colors$Group,
    smooth=TRUE)
```

@@

```{r genesVsCentroids, fig.height=6, fig.width=6}
cluster_of_interest = 2
cluster2Genes = names(
    kmeans_clusters$clusters[kmeans_clusters$clusters==cluster_of_interest])
dataToPlot=rbind(y[cluster2Genes[3:6], ],"Centroid"=kmeans_clusters$centroids[cluster_of_interest,])
plot_splines_data(moanin_model, data=dataToPlot, 
    colors=ann_colors$Group, smooth=TRUE,simpleY =FALSE,
    mar=c(1.5,2.5,2,0.1))
```

@@

```{r assigning_scores_and_labels}
# Then assign scores and labels to all the data, using a goodness-of-fit
# scoring function.
scores_and_labels = splines_kmeans_score_and_label(
    moanin_model, kmeans_clusters)
```
